Boosting the Accuracy of Finite Difference Schemes via Optimal Time Step Selection and Non-Iterative Defect Correction

### 2011-07-27 papers/OTSPDE/OTSPDE.tex
Additional revisions to OTSPDE manuscript.

\begin{abstract}
In this article, we present a unified analysis of the simple technique for
boosting the order of accuracy of finite difference schemes for time dependent
partial differential equations by optimally selecting the time step used to
advance the numerical solution and adding defect correction terms in a
non-iterative manner.
The power of the technique is its ability to increase the accuracy of formally
low-order finite difference schemes without major modification to the basic 
numerical algorithm.  Through straightforward numerical analysis arguments, we
explain the origin of the boost in accuracy and estimate the computational cost
of the resulting numerical method.  We demonstrate the utility of optimal time
step (OTS) selection combined with non-iterative defect correction (NIDC) on
several different types of finite difference schemes for a wide array of
classical linear and semilinear PDEs in one and more space dimensions on both
regular and irregular domains.

@@

In this article, we present a unified analysis of the simple technique for
boosting the \emph{order} of accuracy of finite difference schemes for time
dependent PDEs by optimally choosing the time step and, when necessary, adding
defect correction terms in a \emph{non-iterative} manner.  This technique is a
systematic generalization of the ideas underlying the previously mentioned
examples and specific examples that are scattered throughout the literature.
Optimal time step (OTS) selection is based on the observation that a carefully
chosen time step can \emph{simultaneously} eliminate low-order terms in both
the spatial and temporal the discretization errors.  At its essence, OTS
selection combines the powerful idea of optimally choosing numerical parameters
with the approach of using knowledge of the structure of the PDE to reduce
numerical error.  The end result is a strategy for boosting the order of
accuracy of algorithmically simple and computationally low-cost numerical
schemes.

@@

\subsubsection*{Stability Considerations}
We have not yet considered the issue of numerical stability of finite
difference schemes when the time step is set to $\dto$.  For explicit time 
integration schemes, OTS-NIDC is ineffective when the optimal time step 
is greater than the smallest stable time step.  In this situation, it is 
necessary to replace the finite difference approximation of the PDE.  
For example, if the diffusion equation is solved using a 
finite difference scheme based on forward Euler time integration and a 
discretization of the Laplacian with an isotropic leading-order truncation 
error (see Section~\ref{sec:diffusion_eqn_2d}), the optimal time step becomes 
unstable when the dimension of space is greater than 3.  To make OTS-NIDC 
useful for diffusion problems in 4 and more space dimensions, we can switch 
to the DuFort-Frankel scheme, which is explicit and unconditionally 
stable \cite{gko_book}.

@@

In Figure~\ref{fig:4th_order_parabolic_eqn_1d_perf}, observe that the 
2nd-order Crank-Nicholson scheme actually outperforms the OTS-NIDC scheme over 
the high (but still practically useful) range of errors even though 
Table~\ref{tab:comp_perf_vs_err_4th_order_parabolic} predicts that the OTS-NIDC
scheme should do better asymptotically.  The origin of this discrepancy is the
constant hidden by the big-$O$ notation.  In particular, the factor of $7/120$
in the optimal time step for the forward Euler scheme with a fourth-order 
approximation of the bilaplacian leads to a large hidden constant 
for the compute time. 
However, as Figures~\ref{fig:4th_order_parabolic_eqn_1d_perf} and
\ref{fig:4th_order_parabolic_eqn_1d_error_roundoff} show, there is an important 
limitation of the 2nd-order Crank-Nicholson scheme even if we ignore the 
cross-over which will occur at sufficiently small values of the error -- 
round-off error associated with matrix inversion limits the accuracy that is 
achievable by the scheme.  This limitation also exists for the 4th-order 
Crank-Nicholson scheme.

>>>>

\begin{abstract}
In this article, we present a unified analysis of the simple technique for
boosting the order of accuracy of finite difference schemes for time dependent
partial differential equations (PDEs) by optimally selecting the time step
used to advance the numerical solution and adding defect correction terms in a
non-iterative manner.
The power of the technique, which is applicable to time dependent, semilinear,
scalar PDEs where the leading-order spatial derivative has a constant
coefficient, is its ability to increase the accuracy of formally
low-order finite difference schemes without major modification to the basic 
numerical algorithm.  Through straightforward numerical analysis arguments, we
explain the origin of the boost in accuracy and estimate the computational cost
of the resulting numerical method.  We demonstrate the utility of optimal time
step (OTS) selection combined with non-iterative defect correction (NIDC) on
several different types of finite difference schemes for a wide array of
classical linear and semilinear PDEs in one and more space dimensions on both
regular and irregular domains.

@@

In this article, we present a unified analysis of the simple technique for
boosting the \emph{order} of accuracy of finite difference schemes for time
dependent PDEs by optimally choosing the time step and, when necessary, adding
defect correction terms in a \emph{non-iterative} manner.  This technique,
which is applicable when the leading-order spatial derivative in a semilinear,
scalar PDE has a constant coefficient, is a systematic generalization of the
ideas underlying the previously mentioned examples and specific examples that
are scattered throughout the literature.
Optimal time step (OTS) selection is based on the observation that a carefully
chosen time step can \emph{simultaneously} eliminate low-order terms in both
the spatial and temporal discretization errors.  At its essence, OTS
selection combines the powerful idea of optimally choosing numerical parameters
with the approach of using knowledge of the structure of the PDE to reduce
numerical error.  The end result is a strategy for boosting the order of
accuracy of algorithmically simple and computationally low-cost numerical
schemes.

@@

\subsubsection*{Stability Considerations}
We have not yet considered the issue of numerical stability of finite
difference schemes when the time step is set to $\dto$.  For explicit time 
integration schemes, OTS-NIDC is ineffective when the optimal time step 
is greater than the smallest stable time step.  In this situation, it is 
necessary to replace the finite difference approximation of the PDE.  
For example, if the diffusion equation is solved using a 
finite difference scheme based on forward Euler time integration and a 
discretization of the Laplacian with an isotropic leading-order truncation 
error (see Section~\ref{sec:diffusion_eqn_2d}), the optimal time step becomes 
unstable when the dimension of space is greater than 3.  To make OTS-NIDC 
useful for diffusion problems in four and more space dimensions, we can switch 
to the DuFort-Frankel scheme, which is explicit and unconditionally 
stable \cite{gko_book}.

@@

In Figure~\ref{fig:4th_order_parabolic_eqn_1d_perf}, observe that the 
2nd-order Crank-Nicholson scheme actually outperforms the OTS-NIDC scheme over 
the high (but still practically useful) range of errors even though 
Table~\ref{tab:comp_perf_vs_err_4th_order_parabolic} predicts that the OTS-NIDC
scheme should do better asymptotically.  The origin of this discrepancy is the
constants hidden by the big-$O$ notation.  In particular, the factor of $7/120$
in the optimal time step for the forward Euler scheme with a fourth-order 
approximation of the bilaplacian leads to a large hidden constant 
for the compute time for the OTS-NIDC scheme, and the low cost of inverting
the banded-diagonal biaplacian matrix in 1D leads to a relatively small hidden
constant in the Crank-Nicholson schemes.

As Figures~\ref{fig:4th_order_parabolic_eqn_1d_perf} and
\ref{fig:4th_order_parabolic_eqn_1d_error_roundoff} show, there is an
important advantage to explicit time integration schemes for the fourth-order
parabolic equation -- round-off error associated with matrix inversion does 
\emph{not} limit the achievable accuracy of the numerical solution.  In 
constrast, errors introduced in matrix inversion affects both the 2nd- and
4th-order Crank-Nicholson schemes.
